---
title: "Dilution series analysis"
author: "Vincent Caruso"
date: "August 25, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Set up the environment.
```{r setup}

library(dada2)
library(stringr)
library(Biostrings)
library(tidyverse)
library(ShortRead)

options(tibble.print_min = 10, tibble.print_max = 40)
# options(tibble.width = Inf)

data_path <- "~/thesis/data/dilution"
result_path <- "~/thesis/results/dilution"
ref_path <- "~/thesis/references/zymo_SSU"
analysis_path <- "~/thesis/analysis/dilution"
script_path <- "~/thesis/noisy-microbes/scripts"

source(file.path(script_path, "seq_table_processing.R"))

if (!file_test("-d", analysis_path)) dir.create(analysis_path)

```

##Load and format OTU/SV tables from each method

Load and inspect the tables from each clustering method. Modify tables as necessary so that they all have the following format:

1. Rows are OTUs or SVs
2. The first column is the OTU/SV identifier, and the remaining columns are samples' abundances
3. There are no comment lines (just a header with column names)

After modifying the format of each table, write it back to a tab-separated '.txt' file.
```{r load UCLUST}

# load UCLUST table
uclust_otu_file <- file.path(result_path, "uclust/otu_table.txt")
uclust_seqs_file <- file.path(result_path, "uclust/rep_set/pooled_nochim_rep_set.fasta")
uclust_table <- load_uclust(uclust_otu_file, uclust_seqs_file)
uclust_table
write_tsv(uclust_table %>% select(-sequence), file.path(result_path, "uclust/uclust_table.txt"))

sample_names <- colnames(uclust_table)[c(-1, -length(colnames(uclust_table)))]

# load UPARSE table
uparse_otu_file <- file.path(result_path, "uparse/otu_table.txt")
uparse_seqs_file <- file.path(result_path, "uparse/otus.fa")
uparse_table <- load_uparse(uparse_otu_file, uparse_seqs_file, sample_names = sample_names)
uparse_table
write_tsv(uparse_table %>% select(-sequence), file.path(result_path, "uparse/uparse_table.txt"))

# load UNOISE table
unoise_otu_file <- file.path(result_path, "unoise/zotu_table.txt")
unoise_seqs_file <- file.path(result_path, "unoise/zotus.fa")
unoise_table <- load_unoise(unoise_otu_file, unoise_seqs_file, sample_names = sample_names)
unoise_table
write_tsv(unoise_table %>% select(-sequence), file.path(result_path, "unoise/unoise_table.txt"))

# load MED table
med_otu_file <- file.path(result_path, "med/MATRIX-COUNT.txt")
med_seqs_file <- file.path(result_path, "med/NODE-REPRESENTATIVES.fasta")
med_chimera_file <- file.path(result_path, "med/uchime/NODE-chimeras.fasta")
med_table <- load_med(med_otu_file, med_seqs_file, med_chimera_file)
med_table
write_tsv(med_table %>% select(-sequence), file.path(result_path, "med/med_table.txt"))

# load Deblur table
deblur_otu_file <- file.path(result_path, "deblur/all.txt")
deblur_table <- load_deblur(deblur_otu_file)
deblur_table
write_tsv(deblur_table %>% select(-sequence), file.path(result_path, "deblur/deblur_table.txt"))

# load DADA2 table
dada2_otu_file <- file.path(result_path, "dada2/sv_table.no_chim.txt")
dada2_table <- load_dada2(dada2_otu_file)
dada2_table
write_tsv(dada2_table %>% select(-sequence), file.path(result_path, "dada2/dada2_table.txt"))

```


Read in reference sequences, and compute distances of inferred sequences to the reference sequences for each method.
```{r Levenshtein function}

ref_fastas <- list.files(ref_path, pattern = "_16S.fasta$")
ref_seqs <- readDNAStringSet(file.path(ref_path, ref_fastas)) %>% as.character()
strains <- sapply(str_split(names(ref_seqs), "_16S"), `[`, 1)

dist_tables <- lapply(method_tables, compute_dist_to_ref, ref_seqs)
strain_dist_tables <- lapply(dist_tables, collapse_group_dist, strains)

```


Read in the results of the BLAST searches.
```{r load blast tables}

blast_files <- file.path(analysis_path, paste(methods, "blast.txt", sep = "_"))
blast_tables <- lapply(blast_files, load_blast)
names(blast_tables) <- methods

```


Annotate sequences in sequence tables as "reference" if they exactly match a reference sequence, or as "ref_noisy" if they are within a specified Levenshtein distance range from a reference sequence. 
```{r compute max_dist sensitivity}

max_dist <- seq(1, 20, 1)
class_tables <- list(tibble(), tibble(), tibble(), tibble(), tibble(), tibble())
names(class_tables) <- methods

for (val in max_dist){
  # combine tables in a list
  methods <- c("uclust", "uparse", "unoise", "med", "deblur", "dada2")
  method_tables <- list(uclust_table, uparse_table, unoise_table, med_table, deblur_table, dada2_table)
  names(method_tables) <- methods
  
  # annotate reference and ref_noisy sequences
  method_tables <- mapply(annotate_ref, method_tables, dist_tables, 
                          MoreArgs = list(sample_names = sample_names, max_dist = val),
                          SIMPLIFY = FALSE)
  
  # annotate contaminant and contam_noisy sequences
  method_tables <- mapply(annotate_contam, method_tables, blast_tables, 
                          MoreArgs = list(sample_names = sample_names, max_off = val), SIMPLIFY = FALSE)
  method_tables <- lapply(method_tables, function(mt){
    mt <- mt %>% select(-sequence, sequence)
    return(mt)
  })
  
  # compute a summary of each method
  method_summary <- mapply(summarize_seqs, method_tables, dist_tables, 
                           MoreArgs = list(sample_names = sample_names, strains = strains, max_dist = val),
                           SIMPLIFY = FALSE)
  method_summary <- lapply(method_summary, sanity_check_summary)

  class_tables <- mapply(function(ctab, msum, v){
    new_rows <- msum %>% select(sample, reference, ref_noisy, contaminant, contam_noisy, other, total, check_sum) %>%
      mutate(max_dist = v)
    new_tab <- rbind(ctab, new_rows)
    return(new_tab)
  }, class_tables, method_summary, MoreArgs = list(v = val), SIMPLIFY = FALSE)
  
}

```


```{r plot sensitivity results}

meth <- "uclust"
dt <- class_tables[[meth]]

class_lines <- ggplot(dt, aes(x = max_dist)) +
  # geom_point(aes(color = method), size = 2) +
  # geom_line(aes(y = ref_noisy, group = sample, color = sample), linetype = 1) +
  # geom_line(aes(y = contaminant, group = sample, color = sample), linetype = 3) +
  # geom_line(aes(y = contam_noisy, group = sample, color = sample), linetype = 4) +
  geom_line(aes(y = other, group = sample, color = sample), linetype = 8) +
  # geom_line(aes(y = ref_noisy + other, group = sample, color = sample), linetype = 3) +
  labs(title = "Number of 'ref_noisy' sequences vs. distance from reference", 
       subtitle = "Counts in other classes shown as dotted lines",
       x = "maximum distance from reference", y = "sequence count") +
  # scale_color_brewer(palette = "Dark2") +
  theme(legend.position = "right",
        legend.justification = c(1, 1))
class_lines

```

##Create various plots to visualize results

Reformat annotated sequence tables for plotting with `ggplot`.
```{r prep for ggplot}

seq_stats <- lapply(method_summary, compute_pr_seqs)

# Combine all method tables into one large table, then gather the sample columns into a single column
method_tables_gg <- lapply(method_tables, gather_samples, sample_names, "sample")
method_tables_gg <- lapply(method_tables_gg, annotate_norms, group = "sample")
all_table_gg <- do.call("rbind", method_tables_gg) %>% mutate(method = rep(methods, times = sapply(method_tables_gg, nrow)))

read_stats <- lapply(method_tables_gg, compute_pr_reads, group = "sample")

```


Plot distribution boxplots of non-reference read counts for "contaminant" sequences vs non-contaminant ("other") sequences.
```{r snr contaminant vs other}

mname = "dada2"

#snr_box_contam <- ggplot(dada2_gg[!dada2_gg$reference & !dada2_gg$noisy, ], aes(x = sample, y = log10_norm)) +
snr_box_contam <- ggplot(all_table_gg %>% filter(method == mname, !ref_like), aes(x = sample, y = log10_norm_med)) +
  geom_boxplot(aes(fill = factor(contam_like, levels = c(TRUE, FALSE)))) +
  labs(title = "Distribution of log10(read counts) for non-reference contaminant and non-contaminant sequences",
          subtitle = "Dilution series processed with DADA2", x = "sample dilution", y = "log10(read counts)") +
  scale_fill_manual(name = "sequence identity", labels = c("contaminant-like", "other"), values = c("blue", "orange")) +
  scale_x_discrete(labels = dilution_labels) +
  theme(legend.position = c(0.95, 0.05),
        legend.justification = c(1, 0))
snr_box_contam

```


Plot bar plots of summary stats (number of "reference", "ref_noisy", "contaminant", "other").
```{r summary stat bar plots}

all_summary_gg <- do.call("rbind", method_summary) %>% as.tibble %>% mutate(method = rep(methods, each = length(sample_names)))
all_summary_gg <- gather(all_summary_gg, "statistic", "value", -sample, -method)

ref_mean <- all_table_gg %>% filter(reference) %>% .[["count"]] %>% mean()
ref_noisy_mean <- all_table_gg %>% filter(ref_noisy) %>% .[["count"]] %>% mean()
contam_mean <- all_table_gg %>% filter(contaminant) %>% .[["count"]] %>% mean()
contam_noisy_mean <- all_table_gg %>% filter(contam_noisy) %>% .[["count"]] %>% mean()
other_mean <- all_table_gg %>% filter(other) %>% .[["count"]] %>% mean()

all_summary_gg <- all_summary_gg %>% mutate(norm_median = rep(1, nrow(all_summary_gg)))
all_summary_gg[all_summary_gg$statistic == "reference", "norm_median"] <- all_summary_gg %>% 
  filter(statistic == "reference") %>% .[["value"]] * log10(ref_mean)
all_summary_gg[all_summary_gg$statistic == "noisy", "norm_median"] <- all_summary_gg %>% 
  filter(statistic == "ref_noisy") %>% .[["value"]] * log10(ref_noisy_mean)
all_summary_gg[all_summary_gg$statistic == "contaminant", "norm_median"] <- all_summary_gg %>% 
  filter(statistic == "contaminant") %>% .[["value"]] * log10(contam_mean)
all_summary_gg[all_summary_gg$statistic == "contam_noisy", "norm_median"] <- all_summary_gg %>% 
  filter(statistic == "contam_noisy") %>% .[["value"]] * log10(contam_mean)
all_summary_gg[all_summary_gg$statistic == "other", "norm_median"] <- all_summary_gg %>% 
  filter(statistic == "other") %>% .[["value"]] * log10(other_mean)

bar_colors <- brewer.pal(9, "Set1")

summary_bars <- ggplot(data = all_summary_gg %>% filter(statistic %in% c("reference", "ref_noisy", "contaminant", "contam_noisy", "other")), 
                       aes(x = sample)) +
  geom_col(aes(y = log2(1 + value), fill = statistic), position = position_dodge()) +
  facet_wrap(~method) +
  scale_fill_manual(values = ref_colors[c(1, 3, 4, 6, 8)])
  #scale_fill_brewer(palette = "Set1")
summary_bars

summary_pct_bars <- ggplot(data = all_summary_gg %>% filter(statistic %in% c("pct_ref", "pct_ref_noisy", "pct_contam", "pct_contam_noisy", "pct_other")), 
                       aes(x = sample)) +
  geom_col(aes(y = value, fill = statistic), position = position_dodge()) +
  facet_wrap(~method) +
  scale_fill_manual(values = bar_colors[c(3, 4, 5, 2, 1)])
  #scale_fill_brewer(palette = "Set1")
summary_pct_bars

```